---
title: "Bears"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bears}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gamkapva)
```




* Dennis B., Munholland P.L. & Michael J. Scott (1991).  Estimation of growth and extinction parameters for endangered species. Ecol. Monogr. 61:115-143.

* Dennis B. & Taper M.L. (1994) Density dependence in time series observations of natural populations: estimation and testing. Ecol. Monogr., 64, 205-224.
- LRT test for density dependence
- Species: bear, 2 Yellowstone Elk populations

* Holmes, EE.  2001.  Estimating risks in declining populations with poor data
- Running sums
- Covered in Morris & Doak

* Morris and Doak.  2002.

* Lindley S.T. (2003) Estimation of population growth and extinction parameters from noisy data. Ecol. Appl., 13: 806-813.
- Species: California sea otter and Yellowstone
grizzly bear
- no code?


Dennis and Poncian.  2014,  Density-dependent state-space model for population-abundance
data with unequal time intervals
- bobcats
- R script - custom
- Ornstein-Uhlenbeck state space model

Eberhardt et al 
Table 3
1959 to 1985

```{r}

```


