---
title: "Untitled"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Adapting SAS PROC MIXED code to R lme function to fit state-space models


Stables, Taper & Dennis.  2004. Estimating population trend & process variation for PVA in the presence of sampling error.  Ecology.  https://doi.org/10.1890/03-3101
proc mixed  noitprint noinfo data = in; 
model W= / DDFM=KENWARDROGER;
repeated /type=toep(2) subject=intercept;
estimate 'trend' intercept 1"

Dennis et al.  2006.  Estimating density dependence, proces noise, & observation error  Ecological Monographs.  https://doi.org/10.1890/0012-9615(2006)76[323:EDDPNA]2.0.CO;2
class time;
model y= ;
random time;
repeated / type=ar(1) subject=intercept;
estimate 'intercept' intercept 1;

Staudenmayer and John P. Buonaccorsi.  2006.  Measurement Error in a random walk model with Applications to Population Dynamics.  Biometrics.

Buonaccorsi & Staudenmayer J. (2009) Statistical methods to correct for observation error in a density-independent population model. Ecol. Monogr. 79:  299-324. 


https://doi.org/10.1890/07-1299.1
proc mixed data=b asycov covtest CL;
model yhat = con / noint DDFM=satterth solution cl;
random z1-z&num/type=toep(1);
run;


### Stabples, Taper & Dennis 2004

"The vector W is a series of the observed log population growth rate at each of the (T − 1) one‐step time intervals in the time series of observations."


   Input is yearly population estimates after cards statement  */

dm "log;clear;out;clear;";

data in;
input Nt @@;
retain ntmp 1;
W = log(Nt)-log(ntmp);
ntmp = Nt;
if _n_ = 1 then delete;
cards;   
   38
    51
    46
    52
    53
    28
    34
    36
    19
    23
    28
    22
    13
    13
    19
    12
;

/*  fits intercept only model with toeplitz covariance structure 
    and tests for nonzero trend (intercept value) with Kenward and
    Roger modified degrees of freedom                                  */
proc mixed  noitprint noinfo data = in; 
model W= / DDFM=KENWARDROGER;
repeated /type=toep(2) subject=intercept;
estimate 'trend' intercept 1;



```{r}
condor.N <- c(38,51,46,52,53,28,34,36,19,23,28,22,13,13,19,12)
condor.length <- length(condor.N)
W <- condor.N[-1]/condor.N[-condor.length]
time <- 1:length(W)

library(nlme)

x <- gls(log(W) ~ 1, 
            correlation = corARMA(,~1, p = 2, q = 0))

anova(x)
# SE = sqrt(SD)/N
0.3309129*14
```





### Dennis et al 2006

Dennis density dependence



run;
quit;
Nuggets
https://www.rdocumentation.org/packages/agriTutorial/versions/0.1.4/topics/example4

https://www.stat.ncsu.edu/people/davidian/courses/st790/examples/dental_lme.R

Use glmmPQL?
https://fukamilab.github.io/BIO202/08-B-spatial-regression.html


```{r}
#Data: bird counts
N <-c(18, 10,  9, 14, 17,
      14,  5, 10,  9,  5,
      11, 11,  4,  5,  4,
       8,  2,  3,  9,  2, 
       4,  7,  4,  1,  2,
       4, 11, 11,  9,  6)

# Year 
year.t <- 1:length(N)
group  <- 1
plot(N~year.t, type = "b")

# Make dataframe
bbs.ss <- data.frame(N, year.t,group)
bbs.ss$year.fac <- factor(bbs.ss$year.t)


gls.corCAR1 <- gls(log(N) ~ 1, 
        correlation =  corCAR1(form = ~year.t),
            data = bbs.ss,
            method = "ML")

gls.corExp <- lme(log(N) ~ 1, 
                  random = ~1|year.fac,
           correlation =  corExp(form = ~year.t|year.fac,
                            nugget = T),
            data = bbs.ss,
            method = "ML")

getVarCov(gls.corExp)


mod.corAR1.vs2 <- lme(log(N) ~ 1, 
        correlation =  corCAR1(form = ~year.t| year.fac),
            random = ~1|year.fac,
            data = bbs.ss,
            method = "ML",
        control = lmeControl( #opt = "optim",
                        optimMethod = "L-BFGS-B"))
summary(mod.corAR1.vs2)
summary(mod.corAR1.vs2$modelStruct$corStruct)

VarCorr(mod.corAR1.vs2)
getVarCov(mod.corAR1.vs2)
summary(mod.corAR1.vs2)
# Random effects:
#  Formula: ~1 | year.fac
#         (Intercept)  Residual
# StdDev:   0.6550696 0.2456511
0.6550696^2    # 0.4291162
sqrt(0.6550696)# 0.8093637
0.2456511^2    #0.06034446
sqrt(0.2456511)#0.495632

# Correlation Structure: Continuous AR(1)
#  Formula: ~year.t | year.fac 
#  Parameter estimate(s):
# Phi 
0.2 
cs <- mod.corAR1.vs2$modelStruct$corStruct
coef(cs,unconstrained=FALSE)

# Fixed effects: log(N) ~ 1 
#                Value Std.Error DF  t-value p-value
# (Intercept) 1.824743 0.1299152 30 14.04565       0




library(nlme)

mod.ranef.only <- nlme::lme(log(N) ~ 1, 
            random = ~1|year.fac,
            data = bbs.ss,
            method = "ML")
summary(mod.ranef.only)


mod.corAR1.vs1 <- lme(log(N) ~ 1, 
            correlation =  corCAR1(form = ~year.t|group),
            random = ~1|year.fac,
            data = bbs.ss,
            method = "ML")

# Error:
##"Error in lme.formula(log(N) ~ 1, correlation = corAR1(form = ~year.t |  : 
##  incompatible formulas for groups in 'random' and 'correlation'"

# implies corAR1(form = ~year.t|group)
#  Phi = 0

# corAR1 vs corCAR1
#  "AR" vs
mod.corAR1.vs2 <- lme(log(N) ~ 1, 
            correlation =  corCAR1(form = ~year.t),
            random = ~1|year.fac,
            data = bbs.ss,
            method = "ML")




mod.2 <- lme(log(N) ~ 1, 
            random = ~1|year.fac,
            data = bbs.ss,
            method = "ML")
summary(mod.2)
update(mod.2, correlation = corAR1(form = ~1|group))


coef(mod$modelStruct$corStruct)
corStruc(mod)
summary(mod)
fixef(mod)
var(mod)
covar(mod)
ACF(mod, maxLag = 11)
```


```{r}
update(fm3BW.lme, correlation = corGaus(form = ~ Time))
```


```{r}
mod.1 <- lme(log(N) ~ 1, 
            correlation =  corGaus(form = ~ year.t),
            random = ~1|year.fac,
            data = bbs.ss,
            na.action = na.omit,
            method = "ML")
summary(mod.1)
```


```{r}
n <- 6                                              ## Number of time points
x <- MASS::mvrnorm(mu = rep(0,n),
             Sigma = .7 ^ as.matrix(dist(1:n)) )    ## Simulate the process using the MASS package
y <- x + rnorm(n)   
times <- factor(1:n)
levels(times)
group <- factor(rep(1,n))
dat0 <- data.frame(y,times,group)
glmmTMB(y ~ ar1(times + 0 | group), data=dat0)
```


```{r}
glmmTMB(N ~ ar1(year.t + 1 | group), data=bbs.ss)
```




Stables, Taper & Dennis.  2004. Estimating population trend & process variation for PVA in the presence of sampling error.  Ecology.  https://doi.org/10.1890/03-3101
* density independent
* even time intervals
* Species:  Whooping Crane (Grus americana), grizzly bear (Ursus arctos horribilis), California Condor (Gymnogyps californianus), and Puerto Rican Parrot (Amazona vittata).
* "fits intercept only model with toeplitz covariance structure 
    and tests for nonzero trend (intercept value) with Kenward and
    Roger modified degrees of freedom                                  */
proc mixed  noitprint noinfo data = in; 
model W= / DDFM=KENWARDROGER;
repeated /type=toep(2) subject=intercept;
estimate 'trend' intercept 1"

Dennis et al.  2006.  Estimating density dependence, proces noise, & observation error.  Ecological Monographs.  https://doi.org/10.1890/0012-9615(2006)76[323:EDDPNA]2.0.CO;2
- species: Breeding Bird Survey (BBS)
- PROC MIXED
- density dependence (but fit using PROC mixed?)
class time;
model y= ;
random time;
repeated / type=ar(1) subject=intercept;
estimate 'intercept' intercept 1;

Staudenmayer and John P. Buonaccorsi.  2006.  Measurement Error in a random walk model with Applications to Population Dynamics.  Biometrics.
- computations required to fit this model
can be done in SAS with proc mixed or in R/Splus with
lme() (see the electronic appendix in Staples et al., 2004, at
http://esapubs.org/archive/ecol/E085/025/supp-1.htm,
for instance)

Buonaccorsi & Staudenmayer J. (2009) Statistical methods to correct for observation error in a density-independent population model. Ecol. Monogr. 79:  299-324.  https://doi.org/10.1890/07-1299.1
- density independent
- species: grizzly bears, Whooping Cranes, California Condors, and Puerto Rican Parrots
- PROC mixed
proc mixed data=b asycov covtest CL;
model yhat = con / noint DDFM=satterth solution cl;
random z1-z&num/type=toep(1);
run;
/* some other choices for proc mixed with titles modified */
title ’Unconstrained’;
proc mixed data=b asycov covtest CL nobound;
model yhat = con / noint DDFM=satterth solution cl;
random z1-z&num/type=toep(1);
run;
